---
title: "Introduction into machine learning and analyzes of Breast Cancer Proteomes"
author:
  - name: Mats Slik
    affiliation: Hanze University of Applied Sciences
    address:
    - Zernikeplein 11
    - room D1.12
    email: m.p.slik@st.hanze.nl
abstract:the data were used to assess how the mutations in the DNA are affecting the protein expression landscape in breast cancer. Genes in our DNA are first transcribed into RNA molecules which then are translated into proteins. Changing the information content of DNA has impact on the behavior of the proteome, which is the main functional unit of cells, taking care of cell division, DNA repair, enzymatic reactions and signaling etc. They performed K-means clustering on the protein data to divide the breast cancer patients into sub-types, each having unique protein expression signature. They found that the best clustering was achieved using 3 clusters (original PAM50 gene set yields four different subtypes using RNA data).
my question is are there different ways to categorize subtypes of breast cancer other than the PAM50 method, and define them as benign or malignant?

date: June 15th, 2022
output:
    pdf_document:
      toc: true
      toc_depth: 2
---


# Dataset

### information about the data set and the three give files :

#### About Dataset
**Context:**
This data set contains published iTRAQ proteome profiling of 77 breast cancer samples generated by the Clinical Proteomic Tumor Analysis Consortium (NCI/NIH). It contains expression values for ~12.000 proteins for each sample, with missing values present when a given protein could not be quantified in a given sample.

**Content:**

* **File:** 77cancerproteomesCPTACitraq.csv

  + **RefSeqaccessionnumber:** RefSeq protein ID (each protein has a unique
  ID in a RefSeq database)
  + **gene_symbol:** a symbol unique to each gene (every protein is encoded
  by some gene)
  + **gene_name:** a full name of that gene
  + **Remaining columns:** log2 iTRAQ ratios for each sample (protein
  expression data, most important), three last columns are from healthy
  individuals
* **File:** clinicaldatabreast_cancer.csv

  + **First column** "Complete TCGA ID" is used to match the sample IDs in the main cancer proteomes file (see example script).
  + **All other columns** have self-explanatory names, contain data about the cancer classification of a given sample using different methods. 'PAM50 mRNA' classification is being used in the example script.

* **File:** PAM50_proteins.csv

    * **Contains**  the list of genes and proteins used by the PAM50 classification system. The column RefSeqProteinID contains the protein IDs that can be matched with the IDs in the main protein expression data set.

**Past Research:**
Original research paper: https://www.researchgate.net/publication/303509927_Proteogenomics_connects_somatic_mutations_to_signaling_in_breast_cancer

**Summary:** the data were used to assess how the mutations in the DNA are affecting the protein expression landscape in breast cancer. Genes in our DNA are first transcribed into RNA molecules which then are translated into proteins. Changing the information content of DNA has impact on the behavior of the proteome, which is the main functional unit of cells, taking care of cell division, DNA repair, enzymatic reactions and signaling etc. They performed K-means clustering on the protein data to divide the breast cancer patients into sub-types, each having unique protein expression signature. They found that the best clustering was achieved using 3 clusters (original PAM50 gene set yields four different subtypes using RNA data).
my question is are there different ways to categorize subtypes of breast cancer other than the PAM50 method, and define them as benign or malignant?
```{r}
# packages

```
```{r}
proteomes_data <- read.csv(file = "Data//77_cancer_proteomes_CPTAC_itraq.csv")
clinical_data <- read.csv(file = "Data/clinical_data_breast_cancer.csv")
pam50_protein_data <- read.csv(file = "Data/PAM50_proteins.csv")

# showing succeseful loading of data
head(proteomes_data[1:5], n = 5)
head(clinical_data, n=5)
head(pam50_protein_data, n=5)

# showing structure of dataframe
str(proteomes_data)
str(clinical_data)
str(pam50_protein_data)
```

## codebook
```{r}

codebook1 <- read.csv2("Data/77_cancer_proteomes_CPTAC_codebook.txt")
codebook2 <- read.csv2("Data/clinical_data_breast_cancer_codebook.txt")
```

Type any R code in the chunk, for example:
```{r}

```

