---
title: "Introduction into machine learning and analyzes of Breast Cancer Proteomes"
author:
  - name: Mats Slik
    affiliation: Hanze University of Applied Sciences
    address:
    - Zernikeplein 11
    - room D1.12
    - email: m.p.slik@st.hanze.nl

date: September 15th, 2022
output:
    pdf_document:
      toc: true
      toc_depth: 2
---


# Dataset



## About Dataset

### information about the data set and the three give files :
**Context:**
This data set contains published iTRAQ proteome profiling of 77 breast cancer samples generated by the Clinical Proteomic Tumor Analysis Consortium (NCI/NIH). It contains expression values for ~12.000 proteins for each sample, with missing values present when a given protein could not be quantified in a given sample.

**Content:**

* **File:** 77cancerproteomesCPTACitraq.csv

  + **RefSeqaccessionnumber:** RefSeq protein ID (each protein has a unique
  ID in a RefSeq database)
  + **gene_symbol:** a symbol unique to each gene (every protein is encoded
  by some gene)
  + **gene_name:** a full name of that gene
  + **Remaining columns:** log2 iTRAQ ratios for each sample (protein
  expression data, most important), three last columns are from healthy
  individuals
* **File:** clinicaldatabreast_cancer.csv

  + **First column** "Complete TCGA ID" is used to match the sample IDs in the main cancer proteomes file (see example script).
  + **All other columns** have self-explanatory names, contain data about the cancer classification of a given sample using different methods. 'PAM50 mRNA' classification is being used in the example script.

* **File:** PAM50_proteins.csv

    * **Contains**  the list of genes and proteins used by the PAM50 classification system. The column RefSeqProteinID contains the protein IDs that can be matched with the IDs in the main protein expression data set.

**Past Research:**
Original research paper: https://www.researchgate.net/publication/303509927_Proteogenomics_connects_somatic_mutations_to_signaling_in_breast_cancer

**Summary:** the data were used to assess how the mutations in the DNA are affecting the protein expression landscape in breast cancer. Genes in our DNA are first transcribed into RNA molecules which then are translated into proteins. Changing the information content of DNA has impact on the behavior of the proteome, which is the main functional unit of cells, taking care of cell division, DNA repair, enzymatic reactions and signaling etc. They performed K-means clustering on the protein data to divide the breast cancer patients into sub-types, each having unique protein expression signature. They found that the best clustering was achieved using 3 clusters (original PAM50 gene set yields four different subtypes using RNA data).
my question is are there different ways to categorize subtypes of breast cancer other than the PAM50 method, and define them as benign or malignant?
```{r}
# packages
library(pander)
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(stringr)

```
## Exploratory Data Analysis

loading of the dataframes and showing the successful loading and its dimensions.
note only the first 5 columns of "77_cancer_proteomes_CPTAC_itraq.csv" are shown since after column 4 they are the same type.
```{r}
proteomes_data <- read.csv(file = "Data//77_cancer_proteomes_CPTAC_itraq.csv")
clinical_data <- read.csv(file = "Data/clinical_data_breast_cancer.csv")
pam50_protein_data <- read.csv(file = "Data/PAM50_proteins.csv")

# showing succeseful loading of data

# only showing first 5 columns of proteomes
head(proteomes_data[1:5], n = 5)
head(clinical_data, n=5)
head(pam50_protein_data, n=5)

# showing the structure/dimensions of dataframe
cat("77_cancer_proteomes_CPTAC_itraq [ number of rows:", nrow(proteomes_data), "number of columns:",ncol(proteomes_data),'\n')
cat("clinical_data [ number of rows:", nrow(clinical_data), "number of columns:",ncol(clinical_data),'\n')
cat("pam50_protein_data [ number of rows:", nrow(pam50_protein_data), "number of columns:",ncol(pam50_protein_data),'\n')

```
Checking if the Proteomes data has been correctly read.
```{r}
str(proteomes_data)
```
Nothing strange about the Proteomes dat everything seems to be read correct.

Checking if the clinical data  has been correctly read.
```{r}
str(clinical_data)
```
Nothing strange about the clinical data everything seems to be read correct.

Checking if the pam50 protein data  has been correctly read.
```{r}
str(pam50_protein_data)
```
Nothing strange about the pam50 protein data everything seems to be read correct.


### codebook

loading of the created codebooks for the three dataframes.
showing also its contents and successful loading
```{r}

cancer_proteomes_CPTAC_codebook <- read.csv2("Data/77_cancer_proteomes_CPTAC_codebook.txt")
clinical_data_codebook <- read.csv2("Data/clinical_data_breast_cancer_codebook.txt")
PAM50_protein_codebook <- read.csv2("Data/PAM50_protein_codebook.txt", sep = ";")


cancer_proteomes_CPTAC_codebook
clinical_data_codebook
PAM50_protein_codebook
```

## Data observation

there are 12553 rows in the data, these are proteins identifiable with a RefSeq ID number and have 86 columns of witch the last 83 are samples(with named with their identifiers and the last three from healthy individuals.
to further use the data i shall reshape it to make the rows samples and each column a protein
```{r}
# first making a data frame with only the numerical data, samples start at column number 4 til the end column number 86
proteomes_data_numerical <- proteomes_data[4:86]

# checking data
head(proteomes_data_numerical,n=5)

```
transposing the created data frame "proteomes_data_numerical", and adding the refseq ID as column name
```{r}
proteomes_data_numerical_transposed <- t(proteomes_data_numerical)
colnames(proteomes_data_numerical_transposed) <- proteomes_data$RefSeq_accession_number

# checking is succesfull (only showing the first 3 columns since there are 12553 columns
cat("proteomes_data_numerical_transposed[ number of rows:", nrow(proteomes_data_numerical_transposed), "number of columns:",ncol(proteomes_data_numerical_transposed),'\n')
head(proteomes_data_numerical_transposed[,1:3], 10)
```

